```{r}
library(tidyverse)
library(here) 
library(plyr)
library(dplyr)
library("ggplot2")
```

### Import data, the CSV file...

```{r}
Litter <- read.csv("C:/Users/nghal/OneDrive - Namibia University of Science and Technology/Documents/PhD work/Data collection/Data/Litter decomposition/Analyses/Litter083124.csv")

```

### convert character columns to factors

```{r}
Litter$Species <- factor(Litter$Species)
Litter$Litter_substrate <- factor(Litter$Litter_substrate)
Litter$microsite <- factor(Litter$microsite)
```

### Transport Loss Corrections

```{r}
Litter_T0 <- Litter %>%  # new df that includes only T0 bags, for transport loss calculations
  filter(Month == 0, na.rm = TRUE)

Litter_T0 <- Litter_T0 %>%      # create new variable, tranport loss (litter loss in T0 bags, g)
  mutate(trsptloss_g = Initial_mass - Lbag_weight)

Litter_T0 <- Litter_T0 %>%      # create new variable, tranport loss (%)
  mutate(trsptloss_pct = trsptloss_g / Initial_mass * 100)
```

### Calculate means for transport losses for each litter substrates/type

```{r}

T0means <- Litter_T0 |>
  group_by(Litter_substrate) |>
  summarize(  N    = sum(!is.na(trsptloss_pct)),
              mean_trsptloss_pct = mean(trsptloss_pct, na.rm=TRUE),
              sd_trsptloss_pct   = sd(trsptloss_pct, na.rm=TRUE),
              se_trsptloss_pct   = sd_trsptloss_pct / sqrt(N)
  )
print(T0means)  
```

### explore transport loss data

```{r}
hist(Litter_T0$trsptloss_pct)
# check for normality of distribution of transport losses
qqnorm(Litter_T0$trsptloss_pct)
qqline(Litter_T0$trsptloss_pct) # visualize the data - if normal the data should fall fairly closely to the line 

```

##HT: Given the oddity of transport loss distribution, I don't see a compelling argument for a transport loss correction

# Calculate ash-free masses and ash-free mass remaining 

### calculate percent ash of the litter for all litterbags

```{r}

### HT: Note that percent ash seems very low for time 0 -- these should be double-checked to increase our confidence
Litter <- Litter |>
  mutate(
    sample_mass = Sample_cruc_mass - Crucible_mass,
    ash_mass = Ash_cruc_mass - Crucible_mass,
    pct_ash = (ash_mass/sample_mass)*100
  )
```

### Calculate mean pct_ash values from the time 0 means for each of two litter substrates

```{r}

pct_ash_means <- Litter |>
   group_by(Species, Litter_substrate, Month) |>
  summarise(
      N    = sum(!is.na(pct_ash)),
        pct_ash = mean(pct_ash, na.rm=TRUE),
         sd_pct_ash   = sd(pct_ash, na.rm=TRUE),
         se_pct_ash   = sd_pct_ash / sqrt(N))



```

##The above code is giving empty column for sd_pct_ash and se_pct_ash.

```{r}
# define new variables for the inital percent ash for grass and shrub litter
grass_initial_pctash <- as.numeric(pct_ash_means[1,4])
shrub_initial_pctash <- as.numeric(pct_ash_means[5,4])
```

```{r}

# calculate ash-free initial and final litter masses
Litter <- Litter |>
  mutate(
    Initial_mass_ash_free = ifelse(Litter_substrate == "Grass", 
                                   Initial_mass * (100 - grass_initial_pctash) / 100, 
                                   ifelse(Litter_substrate == "Shrub", 
                                          Initial_mass * (100 - shrub_initial_pctash) / 100, 
                                          NA)), # use NA for (non-existent) other cases
    Final_mass_ash_free = Lbag_weight * (100 - pct_ash) / 100,
    Pct_mass_remaining_ashfree = Final_mass_ash_free/Initial_mass_ash_free *100
  ) 
```

### ##Here, I wanted to create a single graph for both microsites

##I will have to rework on the script below now that I have to incorporate "Species", this will become two plots in the end, one for each litter_substrate

```{r}

# Define the microsites and colors
microsites_colors <- c("Vegetated_patch" = "blue", "Unvegetated_patch" = "red")

# Define the unique colors for Litter_substrate
litter_colors <- c("Shrub" = "darkgreen", "Grass" = "darkorange2")

# Filter the data for the microsites
filtered_data <- subset(Ash_free_mean, microsite %in% names(microsites_colors))

pd <- position_dodge(0.1) # move error bars .05 to the left and right

# Plot the data
p <- ggplot(filtered_data, aes(x = Month, y = mean_Ash_free_mass_remaining, shape = Litter_substrate, group = interaction(microsite, Litter_substrate))) +
  geom_line(aes(color = Litter_substrate, linetype = microsite), position = pd, linewidth = 1) +  # Increase line size to 1.5
  geom_point(aes(color = Litter_substrate), position = pd, size = 3) +     
  geom_errorbar(aes(ymin = mean_Ash_free_mass_remaining - se_Ash_free_mass_remaining, ymax = mean_Ash_free_mass_remaining + se_Ash_free_mass_remaining, color = Litter_substrate), width = 0.5) +  # Set color aesthetic to Litter_substrate
  scale_color_manual(values = litter_colors) +  # Use defined colors for Litter_substrate
  scale_shape_manual(values = c(15, 16)) +  # Set shapes for Litter_substrate
  scale_linetype_manual(values = c("Vegetated_patch" = "solid", "Unvegetated_patch" = "dotted")) +  # Set linetypes for microsites
  xlab("Time (months)") +
  ylab("Mass Remaining (%)") +
  ggtitle("Vegetated and Unvegetated Patches") +
  theme_bw() +  # Use theme_bw() for a white background
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    legend.title = element_blank(),
    axis.text.x = element_text(hjust = 1, color = "black", size = 11, vjust = 1),
    axis.text.y = element_text(angle = 0, color = "black", size = 11, hjust = 1, vjust = 1)
  ) +
  coord_cartesian(xlim = c(0, 15)) +
  scale_x_continuous(breaks = c(0, 4, 8, 14)) +
  scale_y_continuous(limits = c(30, max(filtered_data$mean_Ash_free_mass_remaining)), breaks = seq(30, max(filtered_data$mean_Ash_free_mass_remaining), by = 10))


```
